Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2020-07-26 paper/paper.tex
adds updates files versions

While R has several build-in data structures, including data frames, it has not got any internal support for reading, processing, or visualizing spatial data.
However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page ????).
Comparing to dedicated GIS software (e.g., QGIS, ArcGIS), both raster and vector models have several implementations in the R language.

The most prominent package allowing for raster data representation is \emph{raster}.
A \emph{raster} successor aimed at the simpler interface and improved performance, \emph{terra}, is now being developed, however, it could take several years for this package to be adopted by users.
Alternatively, \emph{stars} package can be used to read and process raster data.
It is interconnected with the \emph{sf} package, and beside raster support, it provides tools for spatial data cubes.

The \emph{sp} package was a standard for vector data representation for more than 10 years.
As of 2020, more than 500 R packages directly depends or imports \emph{sp}.
However, it is now in a maintenance mode only with a recommended successor - the \emph{sf} package.

@@

\hypertarget{basic-spatial-data}{%
\subsection{Basic spatial data}\label{basic-spatial-data}}

@@

All of the packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the \texttt{plot()} function.
However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data.
The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
This package aims for creating publication-ready maps.
Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library.
R's general-purpose plotting package \emph{ggplot2} together with its extension \emph{ggspatial} also offers some capabilities for plotting spatial data.

>>>>

While R has several build-in data structures, including data frames, it has not got any internal support for reading, processing, or visualizing spatial data.
However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page 10).
Comparing to dedicated GIS software (e.g., QGIS, ArcGIS), both raster and vector models have several implementations in the R language.

The most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
A \emph{raster} successor aimed at the simpler interface and improved performance, \emph{terra}, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
Alternatively, \emph{stars} package can be used to read and process raster data \cite{Pebesma2019}.
It is interconnected with the \emph{sf} package, and beside raster support, it provides tools for spatial data cubes.

The \emph{sp} package was a standard for vector data representation for more than 10 years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 R packages directly depends or imports \emph{sp}.
However, it is now in a maintenance mode only with a recommended successor - the \emph{sf} package.

@@

Nowadays, spatial data for many spatial scales is available from abundance of sources.
Some of them require registration or having appropriate credentials and allows downloading only from the web interface.
However, many of the sources made their data publicly available either as a direct download or through an API connection.
This is used by some R packages allowing for downloading various environmental spatial dataset.

A plethora of land cover products on global, continental, and country scales has been developed, however, most of them cannot be directly downloaded from R.
The exception is the National Land Cover Database (NLCD) data for the USA, which can be accessed through the \emph{FedData} package \cite{R-FedData}.
The \emph{elevatr} package provides a synthesis raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}.
It uses different elevation products for a given area and selected level of details (zoom/resolution).
The \emph{rgbif} gives access to data from the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}.
The \texttt{get\_data()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim version 1.4, while the new in-development package \emph{geodata} provides an access to WorldClim version 2 data.
Additionally, the \emph{rgee package} \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

\hypertarget{spatial-data-processing}{%
\subsection{Spatial data processing}\label{spatial-data-processing}}

As we mentioned in the previous sections, gathered or obtained data can be represented by different data types and stored in various file formats.
However, several potential obstacles to applying this data for ecological analysis still exist.

The main issue for many landscape ecology studies is related to projections of the used data.
It is not only required to have all of the used datasets in the same projection, but also selection of a proper projection is crucial.
Coordinates in spatial data can represent one of many coordinate reference systems.
Two main groups of coordinate reference systems (CRS) exist, namely geographical and projected, each having many members.
In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
However, in most landscape ecology studies, projected CRSs, which use some measurement units (e.g., meters), should be utilized.
Selected projected CRS should be based on projection that keeps important properties of spatial data intact (e.g., does not distort areas, shapes, distances, angles) and be appropriate for a given study area.
Gladly, all of the packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations.

Second possible issue with spatial data is when the obtained data extends over a larger area than the study region.
Then, some spatial preprecessing, such as vector clipping or raster cropping, is needed.
Packages from Section \ref{sec:spatrep} also make these operations possible.
Additionally, they offer many other operations, such as merging or joining of spatial data, extraction of values from one dataset into another, raster resolution changes, or vector data simplifications.

Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.

@@

All of the packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the \texttt{plot()} function.
However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
This package aims for creating publication-ready maps.
Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
R's general-purpose plotting package \emph{ggplot2} together with its extension \emph{ggspatial} also offers some capabilities for plotting spatial data \cite{Wickham2016a,R-ggspatial}.
